\section{Introduction}

\begin{frame}{Learning Objectives}
\begin{itemize}
    \item Define a software service and microservices architecture.
    \item Explain why microservices became important in modern software systems.
    \item Identify the essential characteristics of microservices.
    \item Describe the importance of low coupling and high cohesion.
    \item Explain practical guidelines for deciding the scope of a microservice.
\end{itemize}
\end{frame}


\begin{frame}{Why This Chapter Matters}
\begin{itemize}
    \item Modern software systems are often too large to build and deploy as one giant block.
    \item Microservices offer a way to split a system into independent parts.
    \item This can make change, scaling, and deployment easier.
    \item But it also creates new challenges in communication, coordination, data consistency, and failure handling.
\end{itemize}
\end{frame}



\begin{frame}{Why System Decomposition Matters}
\begin{itemize}
    \item One of the most important software architecture decisions is how a system is decomposed into components.
    \item Good decomposition supports:
    \begin{itemize}
        \item parallel development by different teams,
        \item easier replacement or reuse of components,
        \item scalability and elasticity in cloud environments,
        \item improved reliability through isolation.
    \end{itemize}
    \item In distributed cloud systems, components must be easy to replicate, move, and update.
\end{itemize}
\end{frame}

\begin{frame}{What Is a Software Service?}
\begin{itemize}
    \item A software service is a component that can be accessed remotely over the Internet.
    \item A request is sent to the service, and the service returns a result.
    \item The service exposes a published interface.
    \item Internal implementation details are hidden from the service user.
    \item The manager of a service is called the \textbf{service provider}.
\end{itemize}
\end{frame}

\begin{frame}{Why Services Fit Cloud-Based Software}
\begin{itemize}
    \item Services are especially useful in cloud-based systems because they can be:
    \begin{itemize}
        \item distributed across multiple computers,
        \item replicated to handle higher loads,
        \item updated independently,
        \item moved from one server to another.
    \end{itemize}
    \item Services typically avoid maintaining local internal state.
    \item Persistent information is usually stored in a database or maintained by the service requester.
\end{itemize}
\end{frame}

\begin{frame}{From Service-Oriented Architecture to Microservices}
\begin{itemize}
    \item The idea of service-oriented architecture is not new.
    \item Earlier service systems often relied on heavyweight web-service standards such as:
    \begin{itemize}
        \item SOAP
        \item WSDL
        \item XML-based messaging
    \end{itemize}
    \item These approaches were often complex and created performance overhead.
    \item Modern microservices favor simpler interfaces and lighter communication mechanisms.
\end{itemize}
\end{frame}

\begin{frame}{What Are Microservices?}
\begin{itemize}
    \item Microservices are small, independent services that can be combined to create applications.
    \item Each microservice should focus on a single business capability or closely related responsibility.
    \item A microservice should be modifiable and redeployable without forcing changes to the rest of the system.
    \item Microservices exchange messages to request services and return responses.
\end{itemize}
\end{frame}

\begin{frame}{Essential Characteristics of Microservices}
\begin{itemize}
    \item \textbf{Self-contained:} a service manages its own data and user-facing behavior.
    \item \textbf{Lightweight:} services communicate with low-overhead protocols.
    \item \textbf{Implementation independent:} different services may use different languages or technologies.
    \item \textbf{Independently deployable:} each service can run and be updated on its own.
    \item \textbf{Business-oriented:} services are organized around business needs, not just technical utilities.
\end{itemize}
\end{frame}

\begin{frame}{Core Design Goal: Low Coupling}
\begin{itemize}
    \item \textbf{Coupling} measures how much one component depends on other components.
    \item Low coupling means a service has few relationships with other services.
    \item This is important because:
    \begin{itemize}
        \item services can change more independently,
        \item updates are less likely to break the rest of the system,
        \item systems are easier to scale and maintain.
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Core Design Goal: High Cohesion}
\begin{itemize}
    \item \textbf{Cohesion} measures how closely related the responsibilities inside a component are.
    \item High cohesion means the parts of a service all contribute to one focused purpose.
    \item High cohesion is desirable because:
    \begin{itemize}
        \item the service is easier to understand,
        \item the service is easier to test,
        \item fewer cross-service calls are needed during execution.
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Single Responsibility in Practice}
\begin{itemize}
    \item Microservice design is strongly influenced by the single responsibility principle.
    \item Each service should do one thing well.
    \item However, ``one thing'' should be interpreted carefully.
    \item Responsibilities should be grouped in ways that reduce unnecessary shared databases and excessive service-to-service communication.
    \item The goal is not the smallest possible service.
    \item The goal is the \textbf{right boundary} for an independently useful service.
\end{itemize}
\end{frame}

\begin{frame}{How Big Should a Microservice Be?}
\begin{itemize}
    \item There is no perfect size rule for microservices.
    \item A practical guideline is the \textbf{rule of twos}:
    \begin{itemize}
        \item a microservice should be developable, testable, and deployable in about two weeks or less,
        \item the development team should be small, often no larger than a ``two-pizza team.''
    \end{itemize}
    \item Size should be judged by responsibility, independence, and manageability, not by lines of code.
\end{itemize}
\end{frame}

\begin{frame}{Key Takeaways from the Introduction}
\begin{itemize}
    \item Microservices grew out of service-based software ideas but use simpler, lighter communication styles.
    \item They are designed to support scalability, flexibility, and independent deployment.
    \item Good microservice design aims for:
    \begin{itemize}
        \item low coupling,
        \item high cohesion,
        \item strong business alignment,
        \item practical service boundaries.
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Microservices Communicate by Message Exchange}
\begin{itemize}
    \item Microservices interact by sending requests and receiving responses.
    \item Messages may include:
    \begin{itemize}
        \item administrative information,
        \item a service request,
        \item the data needed to deliver that service.
    \end{itemize}
    \item The response includes the result of the requested service.
    \item This communication model is central to distributed systems design.
\end{itemize}
\end{frame}


\begin{frame}{Looking Ahead}
\centering
\Large Next: Section 6.1 Microservices

\vspace{1em}

\normalsize
We will now look more closely at the detailed characteristics, design tradeoffs, and examples of microservices.
\end{frame}
