\section{Microservices Architecture}

\begin{frame}{Section 6.2: Microservices Architecture}
\begin{itemize}
    \item A microservices architecture is an \textbf{architectural style}, not a fixed layered template.
    \item It is commonly used for web-based client-server systems.
    \item The server side is implemented as a set of interacting microservices.
    \item This style is intended to address key problems in distributed systems design.
\end{itemize}
\end{frame}

\begin{frame}{Microservices Architecture as a Style}
\begin{itemize}
    \item A layered architecture defines a common set of components used across similar applications.
    \item A microservices architecture does something different.
    \item It provides a tried-and-tested \textbf{way of organizing} distributed software.
    \item In web systems, the client interacts with an application whose server-side behavior is split across multiple services.
\end{itemize}
\end{frame}

\begin{frame}{Why Use Microservices Architecture?}
\begin{itemize}
    \item It can improve scalability.
    \item It can improve resilience.
    \item It can support independent deployment of services.
    \item It can allow services to be restarted or updated without bringing down the whole system.
    \item In cloud environments, scaling out by replicating services is often cheaper than scaling up a single large server.
\end{itemize}
\end{frame}

\begin{frame}{Example Architecture: Client, Gateway, and Services}
\begin{itemize}
    \item A common design uses:
    \begin{itemize}
        \item a client application,
        \item an API gateway,
        \item a collection of backend services.
    \end{itemize}
    \item The gateway routes and translates requests from the client into calls to the appropriate microservices.
    \item This lets internal services change without forcing immediate changes to the client app.
\end{itemize}
\end{frame}

\begin{frame}{A Photo-Printing Example}
\begin{itemize}
    \item The book uses a photo-printing system as an example.
    \item Possible services include:
    \begin{itemize}
        \item registration,
        \item authentication,
        \item upload,
        \item payment,
        \item printing,
        \item dispatch.
    \end{itemize}
    \item An API gateway sits between the mobile app and these backend services.
\end{itemize}
\end{frame}

\begin{frame}{Architectural Design Decisions}
\begin{itemize}
    \item In a microservices system, the development teams for services are autonomous.
    \item However, the overall system still has to work as a coordinated whole.
    \item Key architectural questions include:
    \begin{itemize}
        \item How should the system be decomposed?
        \item How should services communicate?
        \item How should data be distributed and shared?
        \item How should the system be coordinated?
        \item How should failures be detected and managed?
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{How Should the System Be Decomposed?}
\begin{itemize}
    \item Too many microservices increase communication overhead and complexity.
    \item Too few microservices make services less autonomous and harder to change independently.
    \item Decomposition should balance:
    \begin{itemize}
        \item performance,
        \item independence,
        \item maintainability,
        \item business capability alignment.
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Guidelines for Decomposition}
\begin{enumerate}
    \item \textbf{Balance fine-grain functionality and performance.}
    \item \textbf{Follow the common closure principle.}
    \item \textbf{Associate services with business capabilities.}
    \item \textbf{Design services to own the data they need.}
\end{enumerate}
\end{frame}

\begin{frame}{Common Closure and Business Capability}
\begin{itemize}
    \item The \textbf{common closure principle} says that elements likely to change together should be located together.
    \item This reduces the chance that a single requirement change forces changes across several services.
    \item Services should also be aligned with \textbf{business capabilities}, not arbitrary technical fragments.
    \item A useful service boundary usually reflects a coherent real-world responsibility.
\end{itemize}
\end{frame}

\begin{frame}{Starting from a Monolith}
\begin{itemize}
    \item A practical way to identify microservices is to begin with an existing monolithic architecture.
    \item Then identify parts of the system that encapsulate distinct responsibilities.
    \item These responsibilities can be extracted into services over time.
    \item In other words, refactoring a monolith can be a realistic path to microservices.
\end{itemize}
\end{frame}

\begin{frame}{Service Communications}
\begin{itemize}
    \item Services communicate by exchanging messages.
    \item Important design choices include:
    \begin{itemize}
        \item synchronous or asynchronous interaction,
        \item direct service-to-service communication or message broker middleware,
        \item the protocol and message structure to use.
    \end{itemize}
    \item These choices affect performance, flexibility, and complexity.
\end{itemize}
\end{frame}

\begin{frame}{Synchronous vs. Asynchronous Communication}
\begin{itemize}
    \item \textbf{Synchronous communication:}
    \begin{itemize}
        \item a service waits for a reply before continuing,
        \item simpler to understand,
        \item easier to develop initially.
    \end{itemize}
    \item \textbf{Asynchronous communication:}
    \begin{itemize}
        \item services do not have to wait for immediate responses,
        \item can improve resilience and decoupling,
        \item often requires more infrastructure and coordination.
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Direct Communication vs. Message Brokers}
\begin{itemize}
    \item With \textbf{direct communication}, one service calls another service directly.
    \item This is often easier to understand and faster to implement.
    \item With a \textbf{message broker}, services communicate through middleware.
    \item Brokers can support buffering, decoupling, and asynchronous behavior.
    \item However, brokers add infrastructure complexity.
\end{itemize}
\end{frame}

\begin{frame}{Communication Protocols}
\begin{itemize}
    \item A message protocol defines:
    \begin{itemize}
        \item how messages are structured,
        \item what information is included,
        \item how senders and receivers interpret the message.
    \end{itemize}
    \item Modern systems often use:
    \begin{itemize}
        \item HTTP
        \item JSON
        \item REST-style interfaces
    \end{itemize}
    \item The protocol should be standard enough that all services can interoperate consistently.
\end{itemize}
\end{frame}

\begin{frame}{Data Distribution and Sharing}
\begin{itemize}
    \item In principle, each microservice should manage its own data.
    \item In reality, complete data independence is difficult.
    \item Different services often need related or overlapping data.
    \item This creates one of the hardest problems in microservices:
    \begin{itemize}
        \item how to share data while preserving autonomy and consistency.
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Data Design Guidelines}
\begin{enumerate}
    \item Isolate data within each service as much as possible.
    \item If sharing is unavoidable, make shared data read-only when possible.
    \item If services are replicated, include a mechanism to keep replicas consistent.
\end{enumerate}
\end{frame}

\begin{frame}{ACID Consistency vs. Microservice Independence}
\begin{itemize}
    \item Traditional multi-tier systems often use a shared database with ACID transactions.
    \item ACID transactions ensure updates are completed as a single consistent unit.
    \item This works well for strong consistency.
    \item However, shared databases reduce microservice independence.
    \item If strict consistency is required, you may need to accept tighter coupling.
\end{itemize}
\end{frame}

\begin{frame}{Eventual Consistency}
\begin{itemize}
    \item In many microservices systems, replicas may not be immediately identical.
    \item Instead, the system aims for \textbf{eventual consistency}.
    \item Updates are recorded, propagated, and applied across replicas over time.
    \item This improves scalability and service autonomy,
    \item but it means the system may be temporarily inconsistent.
\end{itemize}
\end{frame}

\begin{frame}{Dealing with Inconsistency}
\begin{itemize}
    \item Two important data problems are:
    \begin{itemize}
        \item \textbf{dependent data inconsistency} caused by actions or failures in one service,
        \item \textbf{replica inconsistency} when replicated databases temporarily disagree.
    \end{itemize}
    \item A common technique is to use a \textbf{transaction log} of pending updates.
    \item Compensating actions may also be needed when part of a multi-service workflow fails.
\end{itemize}
\end{frame}

\begin{frame}{Service Coordination}
\begin{itemize}
    \item Many business operations involve several services working together in a workflow.
    \item Example:
    \begin{itemize}
        \item user login,
        \item password checks,
        \item lockout rules,
        \item authentication success or failure.
    \end{itemize}
    \item The architectural question becomes:
    \begin{itemize}
        \item who controls the workflow?
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Orchestration}
\begin{itemize}
    \item In \textbf{orchestration}, a central controller manages the workflow.
    \item The controller tells each service when to act.
    \item Advantages:
    \begin{itemize}
        \item easier to understand the overall process,
        \item easier to debug,
        \item easier to identify failures.
    \end{itemize}
    \item Disadvantage:
    \begin{itemize}
        \item adds a central coordinating component.
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Choreography}
\begin{itemize}
    \item In \textbf{choreography}, there is no central controller.
    \item Services observe events and decide for themselves when to act next.
    \item This can reduce central control and loosen coupling.
    \item However:
    \begin{itemize}
        \item workflows are harder to understand,
        \item debugging is harder,
        \item recovery from failure can be harder.
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Failure Management}
\begin{itemize}
    \item Failures are normal in distributed systems.
    \item Even if individual services are reliable, failures become inevitable at scale.
    \item The architecture must be designed to:
    \begin{itemize}
        \item detect failures,
        \item report failures,
        \item recover from failures,
        \item degrade gracefully when recovery is not immediate.
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Common Failure Types}
\begin{itemize}
    \item \textbf{Internal service failures:}
    \begin{itemize}
        \item bugs,
        \item invalid data handling,
        \item broken internal links or logic.
    \end{itemize}
    \item \textbf{External service failures:}
    \begin{itemize}
        \item dependency unavailable,
        \item dependency too slow,
        \item dependency returns an error.
    \end{itemize}
    \item \textbf{Performance failures:}
    \begin{itemize}
        \item degraded responsiveness under load,
        \item overload conditions,
        \item timeout problems.
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Failure Responses}
\begin{itemize}
    \item A simple approach is to use standard status codes and agreed-upon error responses.
    \item Systems may use:
    \begin{itemize}
        \item timeouts,
        \item retries,
        \item fallback behaviors,
        \item logging and monitoring.
    \end{itemize}
    \item The architecture should make failures visible and diagnosable.
\end{itemize}
\end{frame}

\begin{frame}{Circuit Breakers}
\begin{itemize}
    \item A \textbf{circuit breaker} prevents repeated calls to a failing service.
    \item If a service repeatedly times out or fails:
    \begin{itemize}
        \item the breaker marks it as unavailable,
        \item later requests are blocked or rerouted,
        \item the system periodically tests whether the service has recovered.
    \end{itemize}
    \item This prevents one struggling service from dragging the rest of the system into the swamp.
\end{itemize}
\end{frame}

\begin{frame}{Monitoring and Observability}
\begin{itemize}
    \item In practice, microservices need strong monitoring support.
    \item Useful observations include:
    \begin{itemize}
        \item service response times,
        \item bottlenecks,
        \item frequency of failures,
        \item patterns of degraded performance.
    \end{itemize}
    \item Logs and dashboards help system managers understand where improvements are needed.
\end{itemize}
\end{frame}

\begin{frame}{Section 6.2 Recap}
\begin{itemize}
    \item Microservices architecture is a distributed architectural style built from interacting services.
    \item Architects must make decisions about decomposition, communication, data sharing, coordination, and failure handling.
    \item These decisions involve tradeoffs between simplicity, autonomy, consistency, and resilience.
    \item A good architecture is not just a set of services.
    \item It is a set of services that can work together without collapsing into distributed chaos.
\end{itemize}
\end{frame}

\begin{frame}{Looking Ahead}
\centering
\Large Next: Section 6.3 RESTful Services

\vspace{1em}

\normalsize
We next examine REST as a common communication style for service interaction.
\end{frame}
