\section{Microservices in Practice}

\begin{frame}{Microservices in Practice}
\begin{itemize}
    \item Microservices are not just a programming pattern.
    \item They are a way of organizing software, teams, deployment, and system evolution.
    \item Their goal is to make large systems easier to change, scale, and maintain.
    \item But they do this by introducing new architectural and operational complexity.
\end{itemize}
\end{frame}

\begin{frame}{The Big Idea}
\begin{itemize}
    \item A microservices system is built from independent services.
    \item Each service owns a focused responsibility.
    \item Each service communicates with others through lightweight interfaces.
    \item The hope is that independent parts can evolve without breaking the whole system.
\end{itemize}
\end{frame}

\begin{frame}{Why Teams Use Microservices}
\begin{itemize}
    \item To scale parts of a system independently
    \item To let teams work more autonomously
    \item To deploy and update services separately
    \item To improve resilience by isolating failures
    \item To support modern cloud-based software systems
\end{itemize}
\end{frame}

\begin{frame}{What Microservices Demand in Return}
\begin{itemize}
    \item More network communication
    \item More coordination between services
    \item Harder data consistency problems
    \item More deployment and monitoring complexity
    \item More responsibility placed on development teams
\end{itemize}
\end{frame}

\begin{frame}{The Core Design Tension}
\begin{itemize}
    \item Microservices promise independence.
    \item Real systems still need communication, shared workflows, and consistent data.
    \item So the architect's job is always a balancing act:
    \begin{itemize}
        \item independence vs. coordination
        \item flexibility vs. complexity
        \item speed vs. consistency
        \item autonomy vs. control
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{What Good Microservice Design Tries to Achieve}
\begin{itemize}
    \item Low coupling between services
    \item High cohesion within each service
    \item Clear business-oriented boundaries
    \item Lightweight communication
    \item Independent deployment
    \item Sensible ownership of data and responsibility
\end{itemize}
\end{frame}

\begin{frame}{A Practical Warning}
\begin{itemize}
    \item Microservices are not automatically better than monoliths.
    \item A badly designed microservices system can be harder to understand, test, deploy, and debug than a monolithic system.
    \item Splitting a system into services is only helpful when the service boundaries are meaningful.
    \item Small pieces are useful only when they reduce pain rather than multiply it.
\end{itemize}
\end{frame}

\begin{frame}{When Microservices Make Sense}
\begin{itemize}
    \item The system is large and likely to evolve
    \item Different parts of the system change at different rates
    \item Different teams need to work independently
    \item Some services need to scale much more than others
    \item The organization has the tooling and discipline to handle distributed systems
\end{itemize}
\end{frame}

\begin{frame}{When Microservices May Be the Wrong Choice}
\begin{itemize}
    \item The system is still small or simple
    \item The team is small and highly centralized
    \item The deployment environment is immature
    \item Strong consistency is required everywhere
    \item The cost of coordination is greater than the value of independence
\end{itemize}
\end{frame}

\begin{frame}{What Students Should Remember Most}
\begin{enumerate}
    \item A microservice is a focused, independently deployable service.
    \item Communication, coordination, and data management become central design problems.
    \item REST is one common communication style, but not the only one.
    \item Deployment, monitoring, rollback, and DevOps are part of the architecture.
    \item Good microservices are not just small. They are well-bounded.
\end{enumerate}
\end{frame}

\begin{frame}{The Practical Takeaway}
\begin{itemize}
    \item Microservices are a powerful architectural style for building large, evolving systems.
    \item They work best when the system is carefully decomposed and supported by strong engineering practices.
    \item In the real world, success depends just as much on testing, deployment, monitoring, and team discipline as on code design.
    \item Architecture is not just about structure.
    \item It is about making change survivable.
\end{itemize}
\end{frame}

\begin{frame}{Discussion Questions}
\begin{enumerate}
    \item Why might a company choose microservices instead of a monolithic architecture?
    \item What new problems appear when a system is split into many services?
    \item Why are low coupling and high cohesion so important in microservices?
    \item When might a monolithic system actually be the better engineering decision?
\end{enumerate}
\end{frame}

\begin{frame}{Further Reading}
\begin{itemize}
    \item Sommerville, \textit{Software Engineering}
    \item Sam Newman, \textit{Building Microservices}
    \item Martin Fowler on Microservices
\end{itemize}
\end{frame}
