\chapter{Core Story: Black-Market Horse Racing in DeathSpank’s World}

In the smoky back-alleys of Spanktopia, the \emph{Bureau of Questionable Athletics} runs
unsanctioned races featuring dubious “horses”: skeleton steeds, thonged unicorns, and a
mysterious two-coconut contraption named \emph{Sir Clip-Clop}. Bettors whisper in code,
merchants haggle over ethically ambiguous oats, and the odds are set by an enchanted abacus
with a drinking problem.

\section{Playable Loop }
\begin{enumerate}[label=\textbf{Step \arabic*:}, leftmargin=2.2em]
  \item \textbf{Choose a race card:} a set of $n$ entrants from a stable of $N$ creatures.
  \item \textbf{Assign lanes and start order:} who lines up where (order matters).
  \item \textbf{Distribute upgrade points:} illegal mods (glitter horseshoes, turbo thongs) across $k$ stats.
  \item \textbf{Place a bet:} pick winners, exactas, or “chaos trifecta.”
  \item \textbf{Run the race:} outcomes are sampled; payouts depend on combinatorics-based odds.
\end{enumerate}

\section{Entities}
\begin{itemize}
  \item \textbf{Stable:} $N$ entrants. Example archetypes: Bone Pony, Thongicorn, Greased Donkey, Sir Clip-Clop.
  \item \textbf{Race Card Size:} $n$ selected to race.
  \item \textbf{Stats:} Speed, Skullduggery, Stamina, Sparkle ($k=4$ stats).
  \item \textbf{Illegal Mod Points:} $p$ points distributed across the $k$ stats (with or without caps).
\end{itemize}

\section{Constraints (examples we’ll count later)}
\begin{itemize}
  \item Exactly $n$ of $N$ entrants race (order does not matter for selection).
  \item Lanes are a permutation of the chosen $n$ (order matters).
  \item Mod points: stars-and-bars across $k$ stats, optionally with per-stat caps.
\end{itemize}

